---
title: "Male age-specific fertility rates"
subtitle: "Call with Bruno"
toc: false
output: 
        pdf_document:
                extra_dependencies: ["float"]
---


```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, fig.pos = "H", out.extra = "")
# Suppress summarise info
options(dplyr.summarise.inform = FALSE)

## Load packages ---------------------------------------------------------------

## Install/load packages
packages <- c("tidyverse", "ggplot2", "here", "viridis")
for(p in packages){
        if(!require(p,character.only = TRUE)) install.packages(p)
        library(p,character.only = TRUE)
}

```



# Own child method

$$f_{a-x}(t-x) = \frac{\frac{C_{x,a}(t)}{L_x}}{W_a(t)\frac{L^f_{a-x-0.5}}{L^f_{a-0.5}}}$$
where $C_{x,a}(t)$ are own children aged $x$ of women aged $a$ enumerated in a census conducted at time $t$,

$W_a(t)$ are women aged $a$ at $t$,

$L_x$ inflates the numerator to account for the children dying,

similarly $\frac{L^f_{a-x-0.5}}{L^f_{a-0.5}}$ inflates the denominator to account for the mothers dying.

* Perform this computation for males

* $L_x$ comes from period life table in year $t-x$?

* $L^f_x$ comes from period life table in year $t-a$? (might be an issue to get it by ethnic group)

* Correction for unmatched children? (Higher for the case of father?)

* Sum $f_{a-x}$ over reproductive ages in year $t-x$ and get proportionate AFSR, $f_x^{Om}(t-x)$?


```{r period-mx, warning=F, fig.cap="Visualization of formula"}
library(LexisPlotR)
library(ggplot2)
library(tidyverse)

lexis <- lexis_grid(year_start = 1960, year_end = 1996, age_start = 0, age_end = 36) %>% 
        lexis_cohort(cohort = 1960) %>% 
         lexis_lifeline(birth = "1990-01-01", lwd = 1.5) %>% 
        lexis_year(year = 1995)
lexis <- lexis + 
    labs(x = "Year", y = "Age") +
    annotate("text", x = as.Date('1990-01-01', '%Y-%m-%d'), y = 30, 
                 label = "a-x",
                 size = 5,
             col = "red4") +
    annotate("text", x = as.Date('1995-01-01', '%Y-%m-%d'), y = 5, 
                 label = "x",
                 size = 5,
             col = "red4") +
    annotate("text", x = as.Date('1995-01-01', '%Y-%m-%d'), y = 35, 
                 label = "a",
                 size = 5,
             col = "red4") +
        annotate("text", x = as.Date('1995-06-15', '%Y-%m-%d'), y = 20, 
                 label = "t",
                 size = 5,
             col = "red4") +
        theme(axis.text.x = element_text(angle = -45, size = 7))
lexis

```



## Required data

* Number of children in single years by age of their fathers in single years (matched children)

* Number of children not living with their fathers by single years (unmatched children)

* Total number of women by single years

* Level of mortality both sexes and for female



# Schoumaker (2019)

$$TFR^m(t) = \frac{B(t)}{\sum^{79}_{x=15}f_x^{Om}(t) \cdot M_x(t)}$$

$$ASFR_x(t) = TFR^m(t) \cdot f_x^{Om}(t)$$

* Use own-child method to get both $TFR^m(t)$ and $f_x^{Om}(t)$? Or directly to get $f^m_x(t)$?



# Points from call

* IPUMS: American community survey (yearly)

    * Apply own-child method per ethnic group
    
    * Reverse survival on 5 years
    
    * Var $poploc$ for children <5 years old
    
    * 80% live with their father
    
    * 20% have dead father/father living in another HH: need to be redistributed