---
title: "Parental loss from drugs and firearms"
subtitle: "Methodological points"
author: "Ben Schlüter"
date: "2023-06-08"
toc: true
output: 
        pdf_document:
                extra_dependencies: ["float"]
bibliography: refs.bib
---


```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, fig.pos = "H", out.extra = "")
# Suppress summarise info
options(dplyr.summarise.inform = FALSE)

## Load packages ---------------------------------------------------------------

## Install/load packages
packages <- c("tidyverse", "ggplot2", "here", "viridis")
for(p in packages){
        if(!require(p,character.only = TRUE)) install.packages(p)
        library(p,character.only = TRUE)
}

```



# Matrix Kinship model

Kin: parents

-   Cause-specific (drugs, firearms, other) mortality [@Caswell2023]
    -   Track parents' deaths by age and cause
-   Two-sex: male fertility = shifted female fertility [@Schoumaker2019]
-   Time-variant rates (2000-2020)

Model run independently on each ethnic group (Hispanic, Black, White)

* $\boldsymbol{F_t}$ is the $(\omega \cdot \omega)$ fertility matrix, in year $t$. The first row contains the fertility rates at the different ages.

* $\boldsymbol{\tilde{U_t}}$ is a block-matrix containing the survival probabilities and cause-specific probabilities of dying, in year $t$.

* $\omega$ is the highest age group + 1, here 86.



# Boundary conditions of the parents' dynamic

In Figure 1, color code is 

* \textcolor{orange}{boundary condition 1} 

* \textcolor{red}{boundary condition 2}

* \textcolor{darkgray}{dynamic of the age distribution of parents}

```{r bnd-cond, warning=F, fig.cap="Boundary conditions of the parents' dynamic"}
ggplot() +
        ## Arrays
        geom_rect(aes(xmin = 45, xmax = 55, ymin = 45, ymax = 55),
                  color = "black", fill = NA) +
        geom_rect(aes(xmin = 29, xmax = 39, ymin = 29, ymax = 39),
                  color = "black", fill = NA) +
        geom_rect(aes(xmin = 17, xmax = 27, ymin = 17, ymax = 27),
                  color = "black", fill = NA) +
        geom_rect(aes(xmin = 5, xmax = 15, ymin = 5, ymax = 15),
                  color = "black", fill = NA) +
        ## Vector of age distribution of parents
        geom_rect(aes(xmin = c(5,17,29), xmax = c(6,18,30), ymin = c(5,17,29), ymax = c(15, 27, 39)),
                  color = "black", fill = "orange") +
        ## Boundary condition for the year 2000
        geom_rect(aes(xmin = seq(6, 14, 1), xmax = seq(7, 15, 1), ymin = 5, ymax = 15),
                  color = "black", fill = "red") +
        ## Meaning of vector
        geom_rect(aes(xmin = 20, xmax = 21, ymin = 17, ymax = 27),
                  color = "black", fill = "darkgray") +
        annotate("text", x = 30, y = 10, 
                 label = "Vector of age distribution of parents (alive & dead by cause)\n at age x+1 of their child in year t+1",
                 size = 2.5) +
        annotate("text", x = 7.5, y = 16, 
                 label = expression(n[x*paste(",")*t]),
                 size = 2.5) +
        annotate("text", x = 20.5, y = 28, 
                 label = expression(n[x+1*paste(",")*t+1]),
                 size = 2.5) +
        ## Evolution of vector
        geom_curve(aes(x = 7.5, y = 17, xend = 19.5, yend = 29),
                   arrow = arrow(length = unit(0.01, "npc"),
                                 type = "closed"),
                   curvature = -0.4) +
        geom_curve(aes(x = 20.5, y = 22, xend = 25, yend = 14),
                   arrow = arrow(length = unit(0.01, "npc"),
                                 type = "closed"),
                   curvature = 0) +
        geom_rect(aes(xmin = 33, xmax = 34, ymin = 29, ymax = 39),
                  color = "black", fill = "darkgray") +
        annotate("text", x = 33.5, y = 41, 
                 label = expression(n[x+2*paste(",")*t+2]),
                 size = 2.5) +
        geom_curve(aes(x = 21, y = 29, xend = 33, yend = 42),
                   arrow = arrow(length = unit(0.01, "npc"),
                                 type = "closed"),
                   curvature = -0.4) +
        geom_curve(aes(x = 33.5, y = 43, xend = 42, yend = 43),
                   arrow = arrow(length = unit(0.01, "npc"),
                                 type = "closed"),
                   curvature = -0.4) +
        annotate("text", x = 10, y = 27, 
                 label = expression(n[x*paste(",")*t]%.%tilde(U)[paste("t")]),
                 size = 2.5) +
        annotate("text", x = 24, y = 42, 
                 label = expression(n[x+1*paste(",")*t+1]%.%tilde(U)[paste("t+1")]),
                 size = 2.5) +
        annotate("text", x = 37, y = 47, 
                 label = expression(n[x+2*paste(",")*t+2]%.%tilde(U)[paste("t+2")]),
                 size = 2.5) +
        geom_rect(aes(xmin = 52, xmax = 53, ymin = 45, ymax = 55),
                  color = "black", fill = "darkgray") +
        annotate("text", x = 45, y = 58, 
                 label = expression(n[x+(T-1)*paste(",")*T-1]%.%tilde(U)[paste("T-1")]),
                 size = 2.5) +
        geom_curve(aes(x = 43, y = 45, xend = 52, yend = 56),
                   arrow = arrow(length = unit(0.01, "npc"),
                                 type = "closed"),
                   curvature = -0.4) +
        ## From one age to the next in boundary condition
        geom_curve(aes(x = seq(5.5, 13.5, 1), y = 4.5, xend = seq(6.5, 14.5, 1), yend = 4.5),
                   arrow = arrow(length = unit(0.01, "npc"),
                                 type = "closed"),
                   curvature = 2, col = "red") +
        ## Age distribution of parents of offsprings
        geom_curve(aes(x = 8, y = 40, xend = c(5.5, 17.5, 29.5), yend = c(10,22,34)),
                   arrow = arrow(length = unit(0.01, "npc"),
                                 type = "closed"),
                   curvature = 0, col = "orange") +
        annotate("text", x = 5, y = 46, 
                 label = expression(pi[t]*paste(" = ")*frac(F[t]^T*paste("(1,:)")%.%n[t], paste("|| ")*F[t]^T*paste("(1,:)")%.%n[t]*paste(" ||"))),
                 size = 3,
                 col = "orange") +
        ## U_tilde matrix
        annotate("text", x = seq(6.5, 14.5, 2.5), y = 0, 
                 label = rep(expression(tilde(U)[paste("2000")]), 4),
                 size = 2.5,
                 col = "red") +
        ## Dim
        annotate("text", x = c(11.5, 1), y = c(17, 10), 
                 label = c("Child age", "Parents' age"),
                 color = c("cyan", "blue"),
                 size = 3) +
        ## Year dim
        annotate("text", x = c(12, 24, 36, 52), y = c(13, 25, 37, 53), 
                 label = c("year 2000", "year 2001", "year 2002", "year 2020"),
                 color = "green",
                 size = 3) +
        geom_point(aes(x = c(41:44), y = c(41:44))) +
        scale_x_continuous(breaks = seq(0, 60, 5), limits = c(0, 60)) +
        scale_y_continuous(breaks = seq(0, 60, 5), limits = c(0, 60)) +
        theme_minimal() +
        theme(axis.title = element_blank(),
              axis.text = element_blank())
```



## Boundary condition 1

Focusing on parents means that fertility enters the model only through the first boundary condition: the age distribution of parents at the birth of their child (\textcolor{orange}{$\pi_t$}). 

Currently, the kin dynamics consists of a full two-sex model ($\boldsymbol{F^f_t}$ and $\boldsymbol{F^m_t}$). Male fertility rates = shifted female fertility rates. The shift is equal to the difference in the mean age at childbearing between the two sexes. The shifts are modeled with a linear model over the years (data from the UN Demographic Yearbook).\

* Sensitivity analysis: use female fertility rates to compute both female and male age distribution at the birth of their children.



## Boundary condition 2

In order to start the dynamics, we assume that the earliest vital rates have been operating for a long period before the year 2000. This means that in order to obtain the vector of the age distribution of parents at age $x$ of their child in the year 2000, we assume

$$\boldsymbol{n_{x, t=2000}} = \textcolor{red}{\boldsymbol{\tilde{U}}}^x_{\textcolor{red}{t=2000}} \cdot \boldsymbol{\pi_{t=2000}}$$
This assumption might not correctly reflect the age distribution of parents at different age of their child (i.e different cohorts) in the year 2000.

Ideally, we would like to be able to correctly reflect any cohort of parent that could have a child aged less than 18 years old, in the year 2000 or after. Hence, the best would be to be able to start the dynamic in the year 1983. FIGURE?

* From the supplementary materials of @verdery2017projections

    * Use life expectancy for Black and White estimated by @national2016health for the period 1980-1999 and convert these into survival probabilities using the Coale-Demeny Model West Life Tables (or modeling methods). This leads to focus only on two ethnic groups.
    
    * Found birth data from CDC starting in 1950.
    
    * Need population counts by ethnic group from the year 1983



# Number of children losing a parent



# Uncertainty in estimates

Monte-Carlo simulation using Chiang method and do similarly as in UN Bayesian population projection



# Parents are shared: overestimation?




# Mx of Fx having the biggest impact for Black: K-K decomposition



# References


<div id="refs"></div>





