---
title: "Parental loss from drugs and firearms"
subtitle: "A cause-specific, two sex, and time-varying matrix kinship model"
author: "Ben Schlüter"
date: "2023-05-15"
toc: true
output: 
        pdf_document:
                extra_dependencies: ["float"]
bibliography: refs.bib
---


```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, fig.pos = "H", out.extra = "")
# Suppress summarise info
options(dplyr.summarise.inform = FALSE)

## Load packages ---------------------------------------------------------------

## Install/load packages
packages <- c("tidyverse", "ggplot2", "here", "viridis")
for(p in packages){
        if(!require(p,character.only = TRUE)) install.packages(p)
        library(p,character.only = TRUE)
}



## Load data -------------------------------------------------------------------

lt.US <- readRDS(here("data_private", "lt_US.rda"))
fx.US <- readRDS(here("data", "fx_US.rda"))
mac.US <- readRDS(here("data", "mac_US.rda"))


## Mortality data from Matt
df.dth <- readRDS(here("data_private",  
                       "state_year_age-sex-race_drug-opioid-mortality.RDS")) %>% 
        rename("age" = age_years) %>% 
        dplyr::select(!c(st_fips, division))

## Kin dynamics
df.prop.dth <- readRDS(here("data_private", "df_prop_dth.rda"))
df.prob.ber.parents <- readRDS(here("data_private", "df_prob_ber_parents.rda"))
df.prob.ber <- readRDS(here("data_private", "df_prob_ber.rda"))


```

# Exploratory data analysis

## Mortality

```{r mx-all, warning = FALSE, fig.cap = "Mortality rates all causes"}
## MX all-causes 
lt.US %>% 
        ggplot(aes(x = age, y = mx, group = year)) +
        facet_grid(sex ~ race_eth) +
        geom_line(aes(col = year)) +
        theme_bw() +
        scale_y_log10() +
        labs(y = "mx (log-scale)")

```

```{r mx-other, warning = FALSE, fig.cap = "Mortality rates all causes except drugs and firearms"}
## Mx all-causes other than drugs & firearms
lt.US %>% 
        mutate(mx_other = (n_deaths - (n_drug + n_firearm))/pop) %>% 
        ggplot(aes(x = age, y = mx_other, group = year)) +
        facet_grid(sex ~ race_eth) +
        geom_line(aes(col = year)) +
        theme_bw() +
        scale_y_log10() +
        labs(y = "mx (log-scale)")

```

```{r mx-drugs, warning = FALSE, fig.cap = "Mortality rates from drugs"}
## Mx drugs
lt.US %>% 
        mutate(mx_drug = n_drug/pop) %>% 
        ggplot(aes(x = age, y = mx_drug, group = year)) +
        facet_grid(sex ~ race_eth) +
        geom_line(aes(col = year)) +
        theme_bw() +
        labs(y = "mx drugs")

```

```{r mx-firearms, warning = FALSE, fig.cap = "Mortality rates from firearms"}
## Mx firearms
lt.US %>% 
        mutate(mx_firearm = n_firearm/pop) %>% 
        ggplot(aes(x = age, y = mx_firearm, group = year)) +
        facet_grid(sex ~ race_eth) +
        geom_line(aes(col = year)) +
        theme_bw() +
        labs(y = "mx firearms")

```

* Stochasticity in $m_x$ for Aian suggests to drop this ethnic group from the analysis.

```{r e0, warning = FALSE, fig.cap = "Life expectancy at birth"}
## Life expectancy at birth
lt.US %>% 
        filter(age == 0) %>% 
        ggplot(aes(x = year, y = ex, group = race_eth, col = race_eth)) +
        facet_wrap(~ sex) +
        geom_line(linewidth = 1) +
        geom_line(aes(y = ifelse(race_eth == "total", ex, NA)), linewidth = 2) +
        theme_bw() +
        scale_color_manual(values = c("aian" = "#440154FF",
                                      "api" = "#3B528BFF",
                                      "black" = "#21908CFF",
                                      "hispanic" = "#5DC863FF",
                                      "white" = "#FDE725FF",
                                      "total" = "#000000")) +
        labs(y = expression(e^0),
             x = "Year",
             col = "Race/ethnic group")

```


## Fertility

```{r mac, warning = FALSE, fig.cap = "Modeled vs observed difference in mean age at childbearing"}
## Modelled difference in MAC
## vs observed MAC
ggplot(mac.US) +
        geom_point(aes(x = year, y = diff.mac)) +
        geom_line(aes(x = year, y = fit, group = 1)) +
        theme_minimal() +
        labs(y = "Difference in mean age at childbearing",
             x = "Year")

```

* Male $f_x$ are obtained by shifting female $f_x$ for each ethnic group. Currently, the shift equals the difference in mean age at childbearing (MAC) at the national level and is allowed to vary over the years. Only five years were available from the UN Demographic Yearbooks. The pattern was pretty linear so I got a shift for all years by fitting a linear model (plotted line) on the available years (plotted points). The decrease in the difference in MAC is in line with @Schoumaker2019.



# Results

* Results focus on Black, White, and Hispanic.

    * High stochasticity in $m_x$ by cause for Aian
    * No fertility rates in the year 2020 for Api and important stochasticity in $m_x$ by cause

```{r prop, warning = FALSE, fig.cap = "Proportion of parents' death"}
## Proportional parents deaths by cause
df.prop.dth %>% 
        filter(cause != "other") %>% 
        ggplot(aes(x = year, y = prop, group = race)) +
        facet_grid(cause ~ focal.age) +
        geom_line(aes(col = race), linewidth = 1) +
        theme_bw() +
        theme(axis.text.x = element_text(angle = 45, 
                                         hjust = 0.7)) +
        scale_color_manual(values = c("aian" = "#440154FF",
                                      "api" = "#3B528BFF",
                                      "black" = "#21908CFF",
                                      "hispanic" = "#5DC863FF",
                                      "white" = "#FDE725FF",
                                      "total" = "#000000")) +
        labs(y = "Proportion of parents' death",
             x = "Year",
             col = "Race/ethnic group")

```


```{r prob-log, warning = FALSE, fig.cap = "Probability of losing one or more parent (log-scale)"}
## Prob of bereavement
## Sum cause-specific probability of losing a parent
df.prob.ber.parents.1 <- df.prob.ber.parents %>% 
        filter(cause != "other") %>% 
        group_by(focal.age, year, race) %>% 
        summarise(prob = sum(prob)) %>% 
        ungroup() %>% 
        mutate(cause = "drugs & firearms")

## Cause-specific probability of losing a parent       
bind_rows(df.prob.ber.parents %>% 
                  filter(cause != "other"),
          ## Add drugs & firearms together
          df.prob.ber.parents.1) %>% 
        mutate(cause = factor(cause,
                              levels = c("drugs", "firearms", "drugs & firearms"))) %>% 
        ggplot(aes(x = year, y = prob, group = race)) +
        facet_grid(cause ~ focal.age) +
        geom_line(aes(col = race), linewidth = 1) +
        theme_bw() +
        theme(axis.text.x = element_text(angle = 45, 
                                         hjust = 0.7)) +
        scale_color_manual(values = c("aian" = "#440154FF",
                                      "api" = "#3B528BFF",
                                      "black" = "#21908CFF",
                                      "hispanic" = "#5DC863FF",
                                      "white" = "#FDE725FF",
                                      "total" = "#000000")) +
        scale_y_log10() +
        labs(y = "Probability of losing one or more parent (log-scale)",
             x = "Year",
             col = "Race/ethnic group")

```

```{r prob, warning = FALSE, fig.cap = "Probability of losing one or more parent"}
## Prob of bereavement
## Cause-specific probability of losing a parent       
bind_rows(df.prob.ber.parents %>% 
                  filter(cause != "other"),
          ## Add drugs & firearms together
          df.prob.ber.parents.1) %>% 
        mutate(cause = factor(cause,
                              levels = c("drugs", "firearms", "drugs & firearms"))) %>% 
        ggplot(aes(x = year, y = prob, group = race)) +
        facet_grid(cause ~ focal.age) +
        geom_line(aes(col = race), linewidth = 1) +
        theme_bw() +
        theme(axis.text.x = element_text(angle = 45, 
                                         hjust = 0.7)) +
        scale_color_manual(values = c("aian" = "#440154FF",
                                      "api" = "#3B528BFF",
                                      "black" = "#21908CFF",
                                      "hispanic" = "#5DC863FF",
                                      "white" = "#FDE725FF",
                                      "total" = "#000000")) +
        labs(y = "Probability of losing one or more parent",
             x = "Year",
             col = "Race/ethnic group")

```


```{r prob-sex, warning = FALSE, fig.cap = "Probability of losing a parent (log-scale)"}
## Prob of bereavement by parent
## Sum cause-specific probability of losing a parent
df.prob.ber.sum <- df.prob.ber %>% 
        filter(cause != "other") %>% 
        group_by(sex, focal.age, year, race) %>% 
        summarise(prob = sum(prob)) %>% 
        ungroup() %>% 
        mutate(cause = "drugs & firearms")

bind_rows(df.prob.ber %>% 
                  filter(cause != "other"),
          ## Add drugs & firearms together
          df.prob.ber.sum) %>% 
        mutate(cause = factor(cause,
                              levels = c("drugs", "firearms", "drugs & firearms"))) %>% 
        ggplot(aes(x = year, y = prob, group = race)) +
        facet_grid(cause ~ focal.age + sex) +
        geom_line(aes(col = race), linewidth = 1) +
        theme_bw() +
        theme(axis.text.x = element_text(angle = 45, 
                                         hjust = 0.7)) +
        scale_color_manual(values = c("aian" = "#440154FF",
                                      "api" = "#3B528BFF",
                                      "black" = "#21908CFF",
                                      "hispanic" = "#5DC863FF",
                                      "white" = "#FDE725FF",
                                      "total" = "#000000")) +
        scale_y_log10() +
        labs(y = "Probability of losing a parent (log-scale)",
             x = "Year",
             col = "Race/ethnic group")

```





# References


<div id="refs"></div>





